Learning by Teaching a Robot:\\The Case of Handwriting

### 2015-12-27 main.tex
Copyedit

\author{Séverin Lemaignan$^1$, Alexis Jacq$^{1,3}$, Deanna Hood$^1$, Fernando Garcia$^1$, \\
    Aude Billard$^2$, Ana Paiva$^3$, Pierre Dillenbourg$^1$ \\
$^1$CHILI Lab, École Polytechnique Fédérale de Lausanne, Suisse,\\
$^2$LASA, École Polytechnique Fédérale de Lausanne, Suisse,\\
$^3$Instituto Superior Técnico, University of Lisbon, Portugal}

@@

Thomas is five and half, and has been diagnosed with visuo-constructive
deficits. He is under the care of an occupational therapist, and tries to
workaround his inability to draw letters in a consistent manner. Vincent is six
and struggles at school with his poor handwriting and even poorer
self-confidence.\footnote{All children's names have been changed.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{henry}
    \caption{\small Thomas teaching Nao how to write numbers, with the help of an
    occupational therapist.}
    \label{fig:henry}
\end{figure}

Remediations for handwriting difficulties traditionally involve long
interventions (at least 10 weeks~\cite{Hoy2011}), essentially consisting in
handwriting training with occupational therapists, and primarily addressing the
\emph{motor} deficits.  Improvements in self-confidence and anxiety occur (at
best) as a side-effect of the child improving their handwriting skills and,
consequently, improving their performance at school.

We present in this article a new take on this educative challenge: a remediation
procedure that involve a ``bad writer'' robot that is \emph{taught} by the
child. By building on the \emph{learning by teaching} paradigm, not only the
child practises handwriting but, as they take on the role of the teacher, they
also positively reinforce their self-esteem and motivation: their social role
shift from the ``underperformer'' to ``the one who knows and teaches''. And by
relying on a robot, we can tailor the exercises and the learning curve to each
children' needs, as we will show in this article.

@@

Robots have been used as teachers or social partners to promote children's
learning in a range of contexts, most commonly related to language
skills~\cite{han2010robot}, and less often to physical skills (such as
calligraphy~\cite{Matsui2013}). Looking at the converse (humans \emph{teaching}
robots), Werfel notes in~\cite{Werfel2014} that most of the work focuses on the
robot's benefits (in terms of language~\cite{Saunders2010} or
physical~\cite{Mulling2013} skills, for example) rather than the learning
experienced by the human tutor themselves.  Our work concentrates on this latter
aspect: by demonstrating handwriting to a robot, we aim at improving the
\emph{child's} performance. Note that our work must be distinguished from
``learning from demonstration'' approaches to robots learning physical skills,
as the agent we present is only simulating fine motor skills for interaction
purposes.

A robotic learning agent which employs the learning by teaching paradigm has
previously been developed by Tanaka and Matsuzoe~\cite{Tanaka2012}. In their
system, children learn vocabulary by teaching the {\sc nao} robot to act out
verbs. The robot is tele-operated (Wizard-Of-Oz) and mimics the actions that the
children teach it, but with no long-term memory or learning algorithm in place.
Our project significantly extends this line of work in two ways. First, by
investigating the context of children's acquisition of a challenging physical
skill (handwriting), and second by proposing a robotic partner which is fully
autonomous in its learning.

@@

We also investigate here a particular role for a robot in the education of
handwriting: not only is the robot actively performing the activity by drawing
letters, but it does so in a way that engages the child in a very specific
social role. The child is the teacher in this relationship and the robot is the
learner: the child is to engage in a (meta-) cognitive relationship with the
robot to try to understand why the robot fails and how to help it best.  Here,
the robot is more than just an activity facilitator or orchestrator -- its
physical presence and embodiment induce agency and anthropomorphising, and
cognitively engage the child into the learning activity (be it consciously or
not).

The commitment of the child into the interaction build on a psychological effect
known as the ``protégé effect''~\cite{Chase2009}: the teacher feels responsible
for his student, commits to the student's success and possibly experiences
student's failure as his own failure to teach. Teachable computer-based agents
have previously been used to encourage this ``protégé effect'', wherein students
invest more effort into learning when it is for a teachable agent than for
themselves~\cite{Chase2009}.  We rely on this cognitive mechanism to reinforce
the child's commitment into the robot-mediated handwriting activity, and we
indeed show sustained child-robot engagement over extended periods of time
(several hours spread of a month).

For these two reasons, our approach is to be distinguished from previous works
in educational robotic. Most of these do not consider the agency induced by the
robot beyond its motivational aspect (playing with an interactive, partially
autonomous device naturally induces some form of anthropomorphizing, which leads
to some level of projected agency, and participate the overall excitement -- at
least, on the short-term, before the novelty effect vanishes).  In our case, the
role of agency is stronger: it induces meta-cognition (``I am interacting with
an agent, so I need to reflect on how to best teach him'') which is beneficial
for the learning process; it also induces a protégé effect (``I want my
robot-agent to succeed!'') which support the commitment of child into the
interaction, also for longer period of time.

Building on these socio-cognitive mechanisms, our intent is therefore to design
a robotic system that would effectively support handwriting remediation in an
original way: by getting children to teach a robotic agent how to write, those
children would both practise without knowing it and recover self-confidence and
self-esteem by supporting a worst-than-themselves robotic ``student''.

The following sections of the paper go into details. We first provides a brief
overview of the robotic system and the interaction it induces. We also present
the machine-learning techniques that allow the robot to learn from the children,
followed by the implementation on a {\sc nao} robot.

We then present and report on the field experiments that we have conducted over
the last two years, including four studies at school, one longer experiment with
eight children at the surgery of an occupational therapist, and two one-month
long case studies. While the focus of the school experiments was mostly the
technical validation and data acquisition, the three other experiments involved
children with actual deficits, and gave us initial insights on the relevance and
effectiveness of our approach.

@@

A tactile tablet (with a custom application) is used for both the robot and the
child to write. During a typical round, the child requests the robot to write
something (a single letter, a number or a full word), and push the tablet
towards the robot. The robot writes on the tablet by gesturing the writing in
the air,  the letters being actually drawn by the tablet application itself. The
child then pull back the tablet, corrects the robot's attempt by writing
him/herself on top or next to the robot's writing (see Figure~\ref{fig:diego}),
and ``send'' his/her demonstration to the robot by pressing a small button on
the tablet. The robot ``learns'' from this demonstration and tries again.

Since the children are assumed to take on the role of the teachers, we had to
ensure they would be able to manage by themselves the turn-taking and the
overall progression of the activity (moving forwards to the next letter or
word). In our design, the turn-taking relies on the robot prompting for feedback
once it is done with its writing (through simple sentences like ``What do you
think?''), and pressing on a small robot icon on the tablet once the child has
finished correcting. In our experiments, and once introduced by the
experimenter, both were easy to grasp for all the children.

@@

Implementing such a system raises several challenges: first, the acquisition,
analyse and learning from hand-written demonstration, which lays at the core of
the our approach, necessitates the development of several algorithms for the
robot to generate initial bad writing and to respond in an adequate manner,
showing visible (but not too quick) writing improvements.

Then, the actual implementation on the robot requires the coordination of
several modules (from performing gestures and acquiring the user's input to the
state machine implementing the high-level behaviour), spread over several
devices (the robot itself, one laptop and up to four tactile tablets for some of
the studies that we conducted). We relied on ROS to ensure the synchronization
and communication between these modules.

@@

    \item we can compute an euclidian distance to the reference letter to assess
        the topological proximity of the demonstration with the expected letter,
        thus providing a quantitative metric of writing performance.

@@

\begin{figure}[ht!]
    \centering
    \subfigure[Nine allographs of the cursive ``h'', next to the reference]{
        \includegraphics[width=0.45\columnwidth]{h}
    }
    \subfigure[Same samples, normalized and projected in the eigenspace spanned from the
        3 first eigenvectors: clusters arise, that actually match writing styles.]{
        \includegraphics[width=0.45\columnwidth]{eigenspace-uniformization}
    }

@@

The algorithm for machine-learning becomes then a simple matter of converging at
a specific pace towards the child demonstration in the eigenspace.
Figure~\ref{learning_6_demos} (p.~\pageref{learning_6_demos}) illustrates the
process with a complete learning cycle of the number ``6''.

@@

Our system is embodied in an Aldebaran's {\sc nao} (V4 or V5, depending on the
studies) humanoid robot. This choice is motivated by its approachable
design~\cite{Gouaillier2008}, its size (58cm) and inherently safe structure
(lightweight plastic) making it suitable for close interaction with children,
its low price (making it closer to what school may afford in the coming years)
and finally its ease of deployment on the field.

Robotic handwriting requires precise closed-loop control of the arm and hand
motion. Because of the limited fine motor skills possible with such an
affordable robot, in addition to the absence of force feedback, we have opted
for \emph{simulated handwriting}: the robot draws letters in the air, and the
actual writing is displayed on a synchronised tablet.

@@

Since the system does not actually require any CPU-intensive process, the laptop
can be removed and the whole logic run on the robot. Due to the relative
difficulty to deploy and debug ROS nodes directly on the robot, the laptop
remains however convenient during the development phase and we kept it during
our experiments.

@@

The system has been deployed and tested in several situations: in three
different schools (more than 70 children aged 5 to 8; relatively short duration
interactions), in the surgery of an occupational therapist (8 children; each
interacting several hours with the system), and during two case studies that
each lasted several weeks. Table~\ref{studies} gives an overview of these
studies.

\begin{table}[ht!]
\centering
\caption{\small Field studies conducted within the project}
\label{studies}
\footnotesize
\begin{tabular}{@{}lp{4cm}p{2.2cm}p{2cm}l@{}}
\toprule
{\bf Study}        & {\bf Type}                               & {\bf Avg. duration}    & {\bf \# Children}                     & {\bf Ages} \\ \midrule
{\it School A (1)} & Unstructured group interaction at school & 16 min/group           & 4 $\times$ 8 children                 & 6-7        \\
{\it School B}     & Individual/Pair interaction at school    & 11 min/group           & 7 (individual) + 7 $\times$ 2 (pairs) & 7-8        \\
{\it School C}     & Pair interaction at school               & 26 min/group           & 7 $\times$ 2                          & 5-6        \\
{\it School A (2)} & Individual interaction at school         & 20 min                 & 6                                     & 5-6        \\ \midrule
{\it Surgery}      & Individual interaction at surgery        & 3 sessions $\times$ 1h & 8                                     & 6-8        \\ \midrule 
{\it Vincent}      & Case-study                               & 4 weeks $\times$ 1.5h  & 1                                     & 6          \\
{\it Thomas}       & Case-study                               & 4 weeks $\times$ 1h    & 1                                     & 5          \\ \bottomrule
\end{tabular}
\end{table}

@@

Critically, these studies were conducted with \emph{whole classes}: we decided
not to select specifically underperforming children as having more children (73
in total) was beneficial for these preliminary studies. Besides, due to ethical
concerns, this would have required complex organization with the school that we
wanted to avoid at the validation stage.

\paragraph{System validation}

From a technical perspective, the system achieves an acceptable level of
reliability and allow a technically sound autonomous interaction. For instance,
during the second school study (\textit{School B}), the robot withstood
interactions which lasted for a total of 160 minutes.  During this time the
robot wrote 335 letters, 152 of which in response to demonstrations received
from the 21 children. Technical intervention was only required for the three
instances that the robot fell during that day.  Otherwise, the technical
components of the system operated autonomously and as expected with all the
groups of children.

@@

The otherwise technical limitations were related to some letters or writing
style (most notably, the ones requiring multiple strokes per letter) not being
adequately processed by the learning algorithm. Support for such letters was
added as a follow-up of the validation studies.

@@

Children's recognition that the robot is writing by itself is critical for our
approach to be effective. When asked, no child indicated that they did not
believe that the robot was writing by itself. There were, at times, questions
about the robot's writing method at the beginning of the interaction, but when
advised that the robot ``tells the tablet what it wants to write,'' this was
accepted by the children.  Besides, teachers interviewed for their feedback on
the system advised that children are asked to draw letters in the air in a
similar manner as part of their handwriting education. The behaviour is hence
not unfamiliar to children.

During two focus groups organized with teachers, the welcomed the approach and
recognized it as useful and promising; this was \textit{a posteriori} confirmed
by multiple spontaneous contacts made by parents and therapists who were looking
forward to use the system with their children.

\paragraph{Sustained engagement}

@@

@@

We also conducted a quantitative assessment of the engagement levels of the
children. Table~\ref{tab:results-with-me-ness} reports the levels of
\emph{with-me-ness} of the children during the second study at school A.
\emph{With-me-ness} is a quantifiable precursor of engagement: it measures the
percentage of time spent by the child focusing onto the task at end. This metric
was first devised in the context of computer-supported
learning~\cite{sharma2014me}, and we have previously studied its applicability
to human-robot interaction and formalized the exact methodology
in~\cite{lemaignan2016realtime}\footnote{Note that with-me-ness is a metric that
allows real-time computation by the robot itself: while we did not yet make
use of it, it does in principle allow the robot to detect on-line possible
disengagements, and eventually address them by adapting its physical
behaviour, switching to different activities, etc.}.

The average with-me-ness is well above 80\%, and confirms that the children
where very much engaged into these 20 minutes of interaction with the robot,
paying close attention to the task.

The other three experiments (the group study at the surgery and the two
case-studies) provide further qualitative evidence of engagement over longer
interaction periods. In particular, as reported hereafter, the two 4-weeks long
case studies that have been conducted so far show that our system can sustain
children' engagement over durations (5 hours) that are getting closer to what is
expected to have an impact in a real therapeutic context.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Surgery Study: How Children Take on the Role of a Teacher}\label{normandie}

@@

The experiment took place at the surgery of an occupational therapist in
Normandy, France. Eight children participated, selected by the occupational
therapist based on their age and type of deficit (all related to handwriting).
Valérie (female, 7 years old), Antoine (male, 6.5) and Johan (male, 7) are under
the direct care of the occupational therapist. Émilien (male, 8) and Mathieu
(male, 7) are repeating their school year because of writing difficulties. Marie
(female, 6) and Adèle (female, 8) are both ranked at the bottom of their
respective classes in writing activities. Nicolas (male, 7) is under the care of
a neurologist, and has been diagnosed with specific language impairment. Given
their age and school year, all those children would be expected to know how to
correctly shape cursive letters.

Over a period of two weeks, each child attended three time a one hour long
session (except for Marie and Adèle who only attended one session). The
experimenter role was limited to the explanation of the task and the basic
tablet usage. For this experiment, the children were provided with two tablets:
one to choose the words (or letters) to teach to the robot, the other one to
write, like in the validation studies. We also provided paper-based templates of
the cursive letters, where the children to ask for them.

We only provided the children with minimal explanations on the task (they would
have to help the robot to improve its writing style), so as to assess if and how
children would naturally take on the role of the teacher. We besides assessed
how seriously they engage into helping the robot through two additional buttons
on the tablet: a green ``thumbs up'' and a red ``thumbs down''. The children
were told to freely use them to evaluate the robot's improvements (``thumb up''
to give positive feedback, ``thumb down'' to convey negative feedback). Our
assessment builds on the hypothesis that the more the child provides feedback to
the robot, the more they assume the role of a teacher. Then, by correlating the
feedback with the actual performance of the robot, we can measure to what extend
the children are seriously assuming their role: if their feedback does correlate
with the actual performance of the robot, the child is likely engaged into a
\emph{serious} teaching role-playing.

@@

All children maintained their engagement during the entire sessions. They
provided on average 42 demonstrations per session. All children did use the
feedback buttons (in total, 99 ``thumb up'' and 33 ``thumb down'' were recorded,
see Table~\ref{table:scores}). This indicates that they are all able to engage
into playing the role of a teacher.

To study the correlation between the children' feedback and the actual
improvements of the robot, we estimate the robot's progress as the difference
between an initial score (average euclidian distance between the shape drawn by
the robot at its first attempt and the shape of the reference letter) and the
robot's score at the current round of demonstrations. We then correlate this
progress to the positive or negative feedback provided by the children (details
of the method are presented in~\cite{jacq2016building}).

We find that the feedback of five out of the eight children do significantly
correlate with the actual performance of the robot (Table~\ref{table:scores}).
This indicates that these children are effectively taking on the teacher's
clothes and are seriously providing feedback to the robot. The observation of
the three remaining children reveal a variety of behaviours (for instance, one
was actually rating how ``nice'' the robot is, and another one had rather
``artistic'' writing style preferences that were independent of the actual
legibility of the robot), and during this specific experiment, we eventually
found that they were all seriously engaged into teaching.

@@

For the first case study, we invited and followed Vincent, a six years old
child, once per week over a period of a month. Our primary aim was to address
the question of whether we can sustain Vincent's engagement and commitment to
the writing activity over such a period.

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{diego}
    \caption{\small Vincent correcting {\sc nao}'s attempt by rewriting the
        whole word. Empty boxes are drawn on the screen to serve as template for the child
        and to make letter segmentation more robust.}
    \label{fig:diego}
\end{figure}

The study took place at our laboratory (Figure~\ref{fig:diego}), and we chose to
design the activity around a storyline meant to be attractive for a 6 years old
boy: one of our {\sc nao} was away for a mysterious scientific mission, and it
needed the support of another one -- which would remain at the lab -- to
interpret curious pictures that were sent every week. Vincent had to help the
second robot understanding the pictures, and since the two robots had somehow
beforehand agreed to communicate with ``letters written like humans'' (\ie
handwritten), Vincent also had to help the robot to write good looking letters
(because, well, this robot was terrible at writing!) The experimental setup was
similar to Figure~\ref{experimental_setup}, except that Vincent had to tell the
robot what to write with small plastic letters (visible behind the robot on
Figure~\ref{fig:diego}).

@@

    \caption{\small Text (in French) generated by the robot, before and after a one
        hour long interaction session with the child. The red box
        highlights one instance of striking improvement of the robot's
        handwriting legibility.}

>>>>

\author{Séverin Lemaignan$^1$, Alexis Jacq$^{1,3}$, Deanna Hood$^1$, Fernando Garcia$^1$, \\
    Aude Billard$^2$, Ana Paiva$^3$, Pierre Dillenbourg$^1$ \\
$^1$CHILI Lab, École Polytechnique Fédérale de Lausanne, Switzerland,\\
$^2$LASA, École Polytechnique Fédérale de Lausanne, Switzerland,\\
$^3$Instituto Superior Técnico, University of Lisbon, Portugal}

@@

Thomas is five and half, and has been diagnosed with visuo-constructive
deficits. He is under the care of an occupational therapist, and tries to
work around his inability to draw letters in a consistent manner. Vincent is six
and struggles at school with his poor handwriting and even poorer
self-confidence.\footnote{All children's names have been changed.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{henry}
    \caption{\small Thomas teaching a \sc{Nao} robot how to write numbers, with 
    the help of an occupational therapist.}
    \label{fig:henry}
\end{figure}

Remediations for handwriting difficulties traditionally involve long
interventions (at least 10 weeks~\cite{Hoy2011}), essentially consisting of
handwriting training with occupational therapists, and primarily addressing the
\emph{motor} deficits.  Improvements in self-confidence and anxiety occur (at
best) as a side-effect of the child improving their handwriting skills and,
consequently, improving their performance at school.

We present in this article a new take on this educative challenge: a remediation
procedure that involves a ``bad writer'' robot that is \emph{taught} by the
child. By building on the \emph{learning by teaching} paradigm, not only does the
child practise handwriting but, as they take on the role of the teacher, they
also positively reinforce their self-esteem and motivation: their social role
shifts from the ``underperformer'' to ``the one who knows and teaches''. And by
relying on a robot, we can tailor the exercises and the learning curve to each
child's needs individually, as we will show in this article.

@@

Robots have been used as teachers or social partners to promote children's
learning in a range of contexts, most commonly related to language
skills~\cite{han2010robot}, and less often to physical skills (such as
calligraphy~\cite{Matsui2013}). Looking at the converse (humans \emph{teaching}
robots), Werfel notes in~\cite{Werfel2014} that most of the work focuses on the
robot's benefits (in terms of language~\cite{Saunders2010} or
physical~\cite{Mulling2013} skills, for example) rather than the learning
experienced by the human tutor themselves.  Our work concentrates on this latter
aspect: by demonstrating handwriting to a robot, we aim at improving the
\emph{child's} performance. Note that our work must be distinguished from
``learning from demonstration'' approaches to robots learning physical skills,
as the agent we present is only simulating poor fine motor skills for interaction
purposes.

A robotic learning agent which employs the learning by teaching paradigm has
previously been developed by Tanaka and Matsuzoe~\cite{Tanaka2012}. In their
system, children learn vocabulary by teaching the {\sc nao} robot to act out
verbs. The robot is tele-operated (Wizard-of-Oz) and mimics the actions that the
children teach it, but with no long-term memory or learning algorithm in place.
Our project significantly extends this line of work in two ways. First, by
investigating the context of children's acquisition of a challenging physical
skill (handwriting), and second by proposing a robotic partner which is fully
autonomous in its learning.

@@

We also investigate here a particular role for a robot in the education of
handwriting: not only is the robot actively performing the activity by drawing
letters, but it does so in a way that engages the child in a very specific
social role. The child is the teacher in this relationship and the robot is the
learner: the child is to engage in a (meta-) cognitive relationship with the
robot to try to understand why the robot fails and how to best help it.  Here,
the robot is more than just an activity facilitator or orchestrator -- its
physical presence and embodiment induce agency and anthropomorphizing, and
cognitively engage the child into the learning activity (be it consciously or
not).

The commitment of the child into the interaction builds on a psychological effect
known as the ``protégé effect''~\cite{Chase2009}: the teacher feels responsible
for his student, commits to the student's success and possibly experiences
student's failure as his own failure to teach. Teachable computer-based agents
have previously been used to encourage this ``protégé effect'', wherein students
invest more effort into learning when it is for a teachable agent than for
themselves~\cite{Chase2009}.  We rely on this cognitive mechanism to reinforce
the child's commitment into the robot-mediated handwriting activity, and we
indeed show sustained child-robot engagement over extended periods of time
(several hours spread over a month).

For these two reasons, our approach is to be distinguished from previous works
in educational robotics. Most of these do not consider the agency induced by the
robot beyond its motivational aspect (playing with an interactive, partially
autonomous device naturally induces some form of anthropomorphizing, which leads
to some level of projected agency, and contributes to the overall excitement -- at
least, on the short-term, before the novelty effect vanishes).  In our case, the
role of agency is stronger: it induces meta-cognition (``I am interacting with
an agent, so I need to reflect on how to best teach him'') which is beneficial
for the learning process; it also induces a protégé effect (``I want my
robot-agent to succeed!'') which supports the commitment of the child into the
interaction, also for longer periods of time.

Building on these socio-cognitive mechanisms, our intent is therefore to design
a robotic system that would effectively support handwriting remediation in an
original way. By getting children to teach a robotic agent how to write, those
children would both practise without knowing it and recover self-confidence and
self-esteem by supporting a worse-than-themselves robotic ``student''.

The following sections of the paper go into the details. We first provide a brief
overview of the robotic system and the interaction it induces. We also present
the machine-learning techniques that allow the robot to learn from the children,
followed by the implementation on a {\sc nao} robot.

We then present and report on the field experiments that we have conducted over
the last two years, including four studies at schools, one longer experiment with
eight children in an occupational therapy clinic, and two one-month
long case studies. While the focus of the school experiments was mostly the
technical validation and data acquisition, the three other experiments involved
children with actual deficits, and gave us initial insights into the relevance and
effectiveness of our approach.

@@

A tactile tablet (with a custom application) is used for both the robot and the
child to write. During a typical round, the child requests the robot to write
something (a single letter, a number or a full word), and pushes the tablet
towards the robot. The robot writes on the tablet by gesturing the writing in
the air,  the letters being actually drawn by the tablet application itself. The
child then pulls back the tablet, corrects the robot's attempt by writing
him/herself on top of or next to the robot's writing (see Figure~\ref{fig:diego}),
and ``sends'' his/her demonstration to the robot by pressing a small button on
the tablet. The robot ``learns'' from this demonstration and tries again.

Since the children are assumed to take on the role of the teachers, we had to
ensure they would be able to manage by themselves the turn-taking and the
overall progression of the activity (moving forwards to the next letter or
word). In our design, the turn-taking relies on the robot prompting for feedback
once it is done with its writing (through simple sentences like ``What do you
think?''), and pressing on a small robot icon on the tablet once the child has
finished correcting. In our experiments, once introduced by the
experimenter, both steps were easy to grasp for the children.

@@

Implementing such a system raises several challenges: first, the acquisition,
analysis and learning from hand-written demonstration, which lays at the core of
the our approach, necessitates the development of several algorithms for the
robot to generate initial bad writing and to respond in an adequate manner,
showing visible (but not too quick) writing improvements.

Then, the actual implementation on the robot requires the coordination of
several modules (from performing gestures and acquiring the user's input to the
state machine implementing the high-level behavior), spread over several
devices (the robot itself, one laptop and up to four tactile tablets for some of
the studies that we conducted). We relied on ROS to ensure the synchronization
and communication between these modules.

@@

    \item we can compute a euclidian distance to the reference letter to assess
        the topological proximity of the demonstration with the expected letter,
        thus providing a quantitative metric of writing performance.

@@

\begin{figure}[ht!]
    \centering
    \subfigure[Nine allographs of the cursive ``h'', next to the reference]{
        \includegraphics[width=0.45\columnwidth]{h}
    }
    \subfigure[The same samples, normalized and projected in the eigenspace spanned by the
        first 3 eigenvectors: clusters arise, that actually match writing styles.]{
        \includegraphics[width=0.45\columnwidth]{eigenspace-uniformization}
    }

@@

The algorithm for machine-learning becomes then a simple matter of converging at
a specific pace towards the child's demonstration in the eigenspace.
Figure~\ref{learning_6_demos} (p.~\pageref{learning_6_demos}) illustrates the
process with a complete learning cycle of the number ``6''.

@@

Our system is embodied in an Aldebaran {\sc nao} (V4 or V5, depending on the
study) humanoid robot. This choice is motivated by its approachable
design~\cite{Gouaillier2008}, its size (58cm) and inherently safe structure
(lightweight plastic) making it suitable for close interaction with children,
its low price (making it closer to what schools may afford in the coming years)
and finally its ease of deployment in the field.

Robotic handwriting requires precise closed-loop control of the arm and hand
motion. Because of the limited fine motor skills possible with such an
affordable robot, in addition to the absence of force feedback, we have opted
for \emph{simulated handwriting}: the robot draws letters in the air, and the
actual writing is displayed on a synchronized tablet.

@@

Since the system does not actually require any CPU-intensive processes, the laptop
can be removed and the whole logic run on the robot. Due to the relative
difficulty to deploy and debug ROS nodes directly on the robot, the laptop
remains however convenient during the development phase and we kept it during
our experiments.

@@

The system has been deployed and tested in several situations: in three
different schools (more than 70 children aged 5 to 8; relatively short duration
interactions), in an occupational therapy clinic (8 children; each
interacting several hours with the system), and during two case studies that
each lasted several weeks. Table~\ref{studies} gives an overview of these
studies.

\begin{table}[ht!]
\centering
\caption{\small Field studies conducted within the project}
\label{studies}
\footnotesize
\begin{tabular}{@{}lp{4cm}p{2.2cm}p{2cm}l@{}}
\toprule
{\bf Study}        & {\bf Type}                               & {\bf Avg. duration}    & {\bf \# Children}                     & {\bf Ages} \\ \midrule
{\it School A (1)} & Unstructured group interaction at school & 16 min/group           & 4 $\times$ 8 children                 & 6-7        \\
{\it School B}     & Individual/Pair interaction at school    & 11 min/group           & 7 (individual) + 7 $\times$ 2 (pairs) & 7-8        \\
{\it School C}     & Pair interaction at school               & 26 min/group           & 7 $\times$ 2                          & 5-6        \\
{\it School A (2)} & Individual interaction at school         & 20 min                 & 6                                     & 5-6        \\ \midrule
{\it Clinic}       & Individual interaction at clinic         & 3 sessions $\times$ 1h & 8                                     & 6-8        \\ \midrule 
{\it Vincent}      & Case-study                               & 4 weeks $\times$ 1.5h  & 1                                     & 6          \\
{\it Thomas}       & Case-study                               & 4 weeks $\times$ 1h    & 1                                     & 5          \\ \bottomrule
\end{tabular}
\end{table}

@@

Critically, these studies were conducted with \emph{whole classes}: we decided
not to select specifically underperforming children as having more children (73
in total) was beneficial for these preliminary studies. In addition, due to ethical
concerns, this would have required complex organization with the school that we
wanted to avoid at the validation stage.

\paragraph{System Validation}

From a technical perspective, the system achieves an acceptable level of
reliability and allows a technically sound autonomous interaction. For instance,
during the second school study (\textit{School B}), the robot withstood
interactions which lasted for a total of 160 minutes.  During this time the
robot wrote 335 letters, 152 of which in response to demonstrations received
from the 21 children. Technical intervention was only required for the three
instances that the robot fell during that day.  Otherwise, the technical
components of the system operated autonomously and as expected with all the
groups of children.

@@

The otherwise technical limitations were related to some letters or writing
styles (most notably, the ones requiring multiple strokes per letter) not being
adequately processed by the learning algorithm. Support for such letters was
added as a follow-up to the validation studies.

@@

Children's recognition that the robot is writing by itself is critical for our
approach to be effective. When asked, no child indicated that they did not
believe that the robot was writing by itself: that it was simply miming the 
actions in the air as the tablet displayed the letters. There were, at times, questions
about the robot's writing without a pen at the beginning of the interaction, but when
advised that the robot ``tells the tablet what it wants to write,'' this was
accepted by the children.  In addition, teachers interviewed for their feedback on
the system advised that children are asked to draw letters in the air in a
similar manner as part of their handwriting education. The behavior is hence
not unfamiliar to children.

During two focus groups organized with teachers, they welcomed the approach and
recognized it as useful and promising; this was \textit{a posteriori} confirmed
by multiple spontaneous contacts made by parents and therapists who were looking
forward to using the system with their children.

\paragraph{Sustained Engagement}

@@

We also conducted a quantitative assessment of the engagement levels of the
children. Table~\ref{tab:results-with-me-ness} reports the levels of
\emph{with-me-ness} of the children during the second study at school A.
\emph{With-me-ness} is a quantifiable precursor of engagement: it measures the
percentage of time spent by the child focusing on the task at hand. This metric
was first devised in the context of computer-supported
learning~\cite{sharma2014me}, and we have previously studied its applicability
to human-robot interaction and formalized the exact methodology
in~\cite{lemaignan2016realtime}\footnote{Note that with-me-ness is a metric that
allows real-time computation by the robot itself: while we did not yet make
use of it, it does in principle allow the robot to detect on-line possible
disengagements, and eventually address them by adapting its physical
behavior, switching to different activities, etc.}.

@@

The average with-me-ness is well above 80\%, and confirms that the children
were very much engaged into these 20 minutes of interaction with the robot,
paying close attention to the task.

The other three experiments (the group study at the clinic and the two
case-studies) provide further qualitative evidence of engagement over longer
interaction periods. In particular, as reported hereafter, the two 4-week long
case studies that have been conducted so far show that our system can sustain
children's engagement over durations (5 hours) that are getting closer to what is
expected to have an impact in a real therapeutic context.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Clinic Study: How Children Take on the Role of a Teacher}\label{normandie}

@@

The experiment took place at an occupational therapy clinic in
Normandy, France. Eight children participated, selected by the occupational
therapist based on their age and type of deficit (all related to handwriting).
Valérie (female, 7 years old), Antoine (male, 6.5) and Johan (male, 7) are under
the direct care of the occupational therapist. Émilien (male, 8) and Mathieu
(male, 7) are repeating their school year because of writing difficulties. Marie
(female, 6) and Adèle (female, 8) are both ranked at the bottom of their
respective classes in writing activities. Nicolas (male, 7) is under the care of
a neurologist, and has been diagnosed with specific language impairment. Given
their age and school year, all of these children would be expected to know how to
correctly shape cursive letters.

Over a period of two weeks, each child attended three times a one hour long
session (except for Marie and Adèle who only attended one session). The
experimenter's role was limited to the explanation of the task and the basic
tablet usage. For this experiment, the children were provided with two tablets:
one to choose the words (or letters) to teach to the robot, the other one to
write, as in the validation studies. We also provided paper-based templates of
the cursive letters, were the children to ask for them.

We only provided the children with minimal explanations on the task (they would
have to help the robot to improve its writing style), so as to assess if and how
children would naturally take on the role of the teacher. We additionally assessed
how seriously they engage into helping the robot through two additional buttons
on the tablet: a green ``thumbs up'' and a red ``thumbs down''. The children
were told to freely use them to evaluate the robot's improvements (``thumb up''
to give positive feedback, ``thumb down'' to convey negative feedback). Our
assessment builds on the hypothesis that the more the child provides feedback to
the robot, the more they assume the role of a teacher. Then, by correlating the
feedback with the actual performance of the robot, we can measure to what extent
the children are seriously assuming their role: if their feedback does correlate
with the actual performance of the robot, the child is likely engaged into a
\emph{serious} teaching role-playing.

@@

All children maintained their engagement during the entirety of the sessions. They
provided on average 42 demonstrations per session. All children used the
feedback buttons (in total, 99 ``thumbs up'' and 33 ``thumbs down'' were recorded,
see Table~\ref{table:scores}). This indicates that they are all able to engage
into playing the role of a teacher.

To study the correlation between the children's feedback and the actual
improvements of the robot, we estimate the robot's progress as the difference
between an initial score (euclidian distance between the shape drawn by
the robot at its first attempt and the shape of the reference letter) and the
robot's score at the current round of demonstrations. We then correlate this
progress to the positive or negative feedback provided by the children (details
of the method are presented in~\cite{jacq2016building}).

We find that the feedback of five out of the eight children does significantly
correlate with the actual performance of the robot (Table~\ref{table:scores}).
This indicates that these children are effectively taking on the teacher's
role and are seriously providing feedback to the robot. The observation of
the three remaining children reveal a variety of behaviors (for instance, one
was actually rating how ``nice'' the robot is, and another one had rather
``artistic'' writing style preferences that were independent of the actual
legibility of the robot), and during this specific experiment, we eventually
found that they were all seriously engaged into teaching.

@@

For the first case study, we invited and followed Vincent, a six year old
child, once per week over a period of a month. Our primary aim was to address
the question of whether we can sustain Vincent's engagement and commitment to
the writing activity over such a period.

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{diego}
    \caption{\small Vincent correcting {\sc nao}'s attempt by rewriting the
        whole word. Empty boxes are drawn on the screen to serve as a template for the child
        and to make letter segmentation more robust.}
    \label{fig:diego}
\end{figure}

The study took place at our laboratory (Figure~\ref{fig:diego}), and we chose to
design the activity around a storyline meant to be attractive for a 6 year old
boy: one of our {\sc nao}s was away for a mysterious scientific mission, and it
needed the support of another one -- which would remain at the lab -- to
interpret curious pictures that were sent every week. Vincent had to help the
second robot understanding the pictures, and since the two robots had somehow
beforehand agreed to communicate with ``letters written like humans'' (\ie
handwritten), Vincent also had to help the robot to write good-looking letters
(because, well, this robot was terrible at writing!). The experimental setup was
similar to Figure~\ref{experimental_setup}, except that Vincent had to tell the
robot what to write with small plastic letters (visible behind the robot on
Figure~\ref{fig:diego}).

@@

    \caption{\small Text (in French) generated by the robot, before and after a one
        hour long interaction session with the child. The red box
        highlights one instance of striking improvement of the robot's
        handwriting legibility \emph{`envoyer'}.}
